---
title: "Lab 2 - Data wrangling"
author: "Fahad Khalid"
date: "Feb 5 2026"
format: 
  html: default # for speed
  pdf: default # for final output

---


```{r}
#| eval: true
#| message: false

library(tidyverse)
```


# Questions

## Part 1

```{r}
midwest |>
  count(state, sort = TRUE)
```

### Question 1
Illinois (IL) has the most at 102, and Indiana (IN) has the second most at 92. Wisconsin (WI) has the fewest at 72.

\newpage

### Question 2
If a county can't repeat in the same state, if a county name appears in the data 5 times, it must be in every state.
```{r}
midwest |>
  count(county) |>
  filter(n == 5)
```

\newpage

### Question 3a
```{r}
midwest |>
  filter(popdensity > 25000) |>
  select(county, state, popdensity, poptotal, area) |>
  arrange(desc(popdensity))
```
### Question 3b
```{r}
midwest |>
  filter(popdensity == max(popdensity)) |>
  select(county, state, popdensity, poptotal, area)
```
\newpage
### Question 4
```{r}
midwest |>
  summarize(
    typical = median(popdensity),
    lower = quantile(popdensity, 0.25),
    upper = quantile(popdensity, 0.75)
  )
```
The distribution of population density of counties is unimodal and extremely right-skewed (which is why we used median for typical population density). A typical Midwestern county has population density of 1156 people per unit area. The middle 50% of the counties have population densities between 622 to 2330 people per unit area.
\newpage

### Question 5
```{r}
midwest |>
  mutate(metro = if_else(inmetro == 1, "Yes", "No")) |>
  count(state, metro) |>
  group_by(state) |>
  mutate(prop = n / sum(n)) |>
  filter(metro == "Yes")
```

\newpage

### Question 6
```{r}
midwest |>
  filter(percollege < 10, percbelowpoverty > 45) |>
  select(county, state, percbelowpoverty, percollege)
```
```{r}
midwest |>
  filter(percollege > 40, percbelowpoverty < 20) |>
  select(county, state, percbelowpoverty, percollege)
```
```{r}

midwest |>
  filter((percollege < 10 & percbelowpoverty > 45) | 
         (percollege > 40 & percbelowpoverty < 20)) |>
  select(county, state, percbelowpoverty, percollege)
```
```{r}
midwest <- midwest |>
  mutate(potential_outlier = if_else(
    (percollege < 10 & percbelowpoverty > 45) | (percollege > 40 & percbelowpoverty < 20),
    "Yes",
    "No"
  ))

midwest |>
  select(county, state, percbelowpoverty, percollege, potential_outlier) |>
  arrange(potential_outlier)
```
```{r}
ggplot(midwest, aes(x = percollege, y = percbelowpoverty, color = potential_outlier)) +
  geom_point() +
  labs(
    x = "% college educated",
    y = "% below poverty",
    title = "Percentage with a college degree versus poverty rate",
    subtitle = "by State in the Midwest",
    color = "Potential Outlier"
  )
```

\newpage

### Question 7
```{r}
state_population <- midwest |>
  group_by(state) |>
  summarize(total_pop = sum(poptotal)) |>
  arrange(desc(total_pop))


```


```{r}
#B) add proportion
state_population |>
  mutate(prop = total_pop / sum(total_pop))
```
IL (Illinois) is the most populous state, containing approximately 27.2% of the Midwest population. WI (Wisconsin) is the least populous, containing approximately 11.6% of the population.
\newpage

### Question 8

\newpage

## Part 2


### Question 9

